---
title: "Favorability of Socialist Candidates Among Latino Voters Who Trace Their Ancestry to Communist Countries"
author: "Jorah Razumeyko"
date: "2024-12-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
library(Rmisc)
library(gt)
library(webshot2)
```
## Abstract:
  This project seeks to determine if a socialist candidate is less popular among Latino Americans who trace their ancestry back to communist countries. We find that such Americans are less favorable of candidates running as "socialist" by looking at the favorability of Bernie Sanders among Latino Americans with different backgrounds.  

## Introduction:
  Some have argued that it is best to avoid running a candidate who accepts the label of "socialist as it may drive down turnout among Latino Americans who disapprove of socialism. This argument assumes that those who trace their ancestry back to Cuba, Venezuela, and Nicaragua, which have had a history of socialist government policies, will be less likely to support a socialist candidate.

## Data:
  We can test this by looking at the [2016 Collaborative Multiracial Post-Election Survey data](https://doi.org/10.3886/ICPSR38040.v2) (CMPS), which asked Latino American respondents "To what country do you or your family trace your ancestry?", resulting in over 3000 responses. Using this we can group together respondents that chose Cuba, Venezuela, and Nicaragua as our "communist-roots" group and all other Latino American respondents as our "non-communist-roots" group. 

  We can pair this with question asking if a respondent has a "favorable view or unfavorable view of Bernie Sanders", who ran as a Democratic Socialist during the Democratic Presidential primary.

```{r, echo = FALSE}
# code to load original data
# #loading in CMPS data

# Full_CMPS <- read.csv("CMPS2016_working.csv")
# nrow(Full_CMPS)

# #subset of just Latino respondents
# Latino_Respondents <- Full_CMPS[!is.na(Full_CMPS$latino_group_identity), ]

# #Smaller data frame with only necessary variables
# truncated <- Latino_Respondents[, c("pid7", "s10", "c2", "c4", "c8")] 

# #grouping together as communist and non communist
# truncated$roots <- "Non-Communist"
# truncated$roots[truncated$s10 %in% c("Cuba", "Venezuela", "Nicaragua")] <- "Communist"
# truncated$roots <- factor(truncated$roots, levels = c("Non-Communist", "Communist"))


# write.csv(truncated, file = "truncated.csv", row.names = FALSE)

```



```{r, echo = FALSE}
#loading in CMPS data


truncated <- read.csv("truncated.csv")



#colors
blue <- rgb(79/255, 148/255, 207/255)
orange <- rgb(255/255, 139/255, 0)

#ancestry data as table 
root_table <- sort(table(truncated$roots), decreasing = TRUE)
perc <- (root_table / sum(root_table)) * 100

#percentage of responses bar plot
barplot(height = perc,
        names.arg = names(root_table),
        col = blue,
        main = "How many Latino voters have roots in Communist Countries\nand how many do not",
        ylab = "Percentage of Responses",
        ylim = c(0, 100)
        
)

```

## Results

Using these two groups we can compare how favorability of Bernie Sanders differs for Latino Americans.

```{r, echo = FALSE}
#excluding respondents unfamiliar with Bernie Sanders and ordering responses

fav <- subset(truncated, c8 != "Not famili")

columns <- c("c8")

# Define the replacements
replacements <- list(
  "Very unfav" = 1,
  "Somewhat u" = 2,
  "Somewhat f" = 3,
  "Very favor" = 4
)

# Apply lapply for recoding
fav[columns] <- as.data.frame(lapply(fav[columns], function(column) {
  recode(column, !!!replacements)
}))


#separating the data into two groups for further analysis:

communist_data <- fav[fav$roots == "Communist", ]

non_communist_data <- fav[fav$roots == "Non-Communist", ]

# count and percentage calculation
communist_table <- table(communist_data$c8)

non_communist_table <- table(non_communist_data$c8)


communist_perc <- (communist_table / sum(communist_table)) * 100
non_communist_perc <- (non_communist_table / sum(non_communist_table)) * 100

# Plot the Communist data
barplot(communist_perc,
        col = orange,
        main = "Majority of Latino voters with roots in Communist Countries\n have a favorable view of Bernie Sanders",
        ylab = "Percentage of Responses",
        ylim = c(0, 40),
        names.arg = c("Very \n Unfavorable",  "Somewhat \n Unfavorable", "Somewhat \n favorable","Very \n favorable" ))


# Plot the Non-Communist data
barplot(non_communist_perc,
        col = blue,
        main = "Strong Majority of Latino voters with roots in Non-Communist \n Countries have a favorable view of Bernie Sanders",
        ylab = "Percentage of Responses",
        ylim = c(0, 40),
        names.arg = c("Very \n Unfavorable",  "Somewhat \n Unfavorable", "Somewhat \n favorable","Very \n favorable" ))


```

This shows that Sanders is more popular among Latinos who do not trace their ancestry to Cuba, Niceragua, and Venezuala. We can test the significance by running a regression


```{r,echo = FALSE}
#Sanders Approval

# Convert to specified levels
fav$roots <- factor(fav$roots, levels = c("Non-Communist", "Communist"))
fav$roots <- relevel(factor(fav$roots), ref = "Non-Communist")

# Fit the linear model using the dummy variables
fav_model <- lm(c8 ~ roots, data = fav)


# create table
coefficients_df <- as.data.frame(summary(fav_model)$coefficients)
coefficients_df <- cbind(Term = rownames(coefficients_df), coefficients_df)
rownames(coefficients_df) <- NULL  # Remove row names

colnames(coefficients_df) <- c("Term", "Estimate", "Std. Error", "t value", "Pr(>|t|)")



gt(coefficients_df) %>%
  tab_header(
    title = "Model Coefficients",
    subtitle = "Summary of Linear Regression Model"
  ) %>%
  fmt_number(
    columns = c("Estimate", "Std. Error", "t value", "Pr(>|t|)"),
    decimals = 3
  )




# Get predictions from the model with confidence intervals
predictions <- predict(fav_model, interval = "confidence")

fav$fit <- predictions[, "fit"]
fav$lwr <- predictions[, "lwr"]
fav$upr <- predictions[, "upr"]


# Create the plot
ggplot(fav, aes(x = roots, y = fit)) +
  geom_point(color = "blue", size = 3) + 
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, color = "black") +  
   ylim(1, 4) +
  labs(
    x = "Countries to which respondents trace their roots to",
    y = "Average Sanders Approval \n (ranging from 1 = Very Unfavorable to 4 = Very favorable)",
    title = "Latinos on average have a favorable view of Sanders, but those \n with roots in Non-Communist countries have a more favorable view"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 10)) 

```


While this data shows a statistically significant effect of having roots in communist countries, it misses the bigger picture that this could simply be a product of Cuban Americans that are more conservative than other Latin Americans. We can get around this by controlling for party affiliation of the respondents. Looking at party identification, going we can see that there much more strong democratic leaning Latino Americans with roots in non-communist countries.


```{r, echo = FALSE}
#Controlling for party affiliation
# count and percentage calculation

fav$pid7 <- as.factor(fav$pid7)
communist_table <- table(communist_data$pid7)


non_communist_table <- table(non_communist_data$pid7)


communist_perc <- (communist_table / sum(communist_table)) * 100
non_communist_perc <- (non_communist_table / sum(non_communist_table)) * 100

fav_model <- lm(c8 ~ pid7, data = fav)


# model for Communist group
fav_communist <- subset(fav, roots == "Communist")
fav_model_communist <- lm(c8 ~ pid7, data = fav_communist)

# model for Non-Communist group
fav_non_communist <- subset(fav, roots == "Non-Communist")
fav_model_non_communist <- lm(c8 ~ pid7, data = fav_non_communist)

# predictions for Communist group
predictions_communist <- predict(fav_model_communist, interval = "confidence")
fav_communist$fit <- predictions_communist[, "fit"]
fav_communist$lwr <- predictions_communist[, "lwr"]
fav_communist$upr <- predictions_communist[, "upr"]

# predictions for Non-Communist group
predictions_non_communist <- predict(fav_model_non_communist, interval = "confidence")
fav_non_communist$fit <- predictions_non_communist[, "fit"]
fav_non_communist$lwr <- predictions_non_communist[, "lwr"]
fav_non_communist$upr <- predictions_non_communist[, "upr"]


# Plot the Communist data
barplot(communist_perc,
        col = orange,
        main = "Party Identification among the Communist Countries group \n is more uniformly distributed between each identification",
        ylab = "Percentage of Responses",
        ylim = c(0, 30),
        names.arg = c("SD", "D", "LD", "I", "LR", "R", "SR"))


# Plot the Non-Communist data
barplot(non_communist_perc,
        col = blue,
        main = "Party Identification among the Non-Communist Countries group is \n skewed left with a considerable number of independents",
        ylab = "Percentage of Responses",
        ylim = c(0, 30),
        names.arg = c("SD", "D", "LD", "I", "LR", "R", "SR"))


# Combine both groups back into one dataset
fav_combined <- rbind(fav_communist, fav_non_communist)

# Create the plot
ggplot(fav_combined, aes(x = pid7, y = fit, color = roots)) +
  geom_point(size = 1,  position = position_dodge(width = 0.4)) +  # Points for predicted values
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, position = position_dodge(width = 0.4)) +
  ylim(1, 4) +  # Adjust y-axis range
  labs(
    x = "Party Identification from Strong Democrat to Strong Republican", 
    y = "Average Sanders Approval \n (ranging from 1 = Very Unfavorable to 4 = Very favorable)", 
    title = "Approval of Sanders does not vary between Communist and non-Communist \n to a statistically significant extent within each party identification group"
  ) +
  scale_x_discrete(labels = c("SD", "D", "LD", "I", "LR", "R", "SR")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 10),
    legend.title = element_blank(),  # Remove legend title
    legend.position = "top"  # Place legend at top
  ) +
  scale_color_manual(values = c("blue", "red"))  # Customize colors for each group





```

If we use party identification data to predict the approal of Sanders, we can see that there is not enough data to determine that they are less popular among Latino Americans with roots in Communist or Socialist countries.

However, it can be useful to compare it to other candidates to see if there is any difference for other politicians. 



```{r, echo = FALSE}
#Obama Approval
favc <- subset(truncated, c2 != "Not famili")

columns <- c("c2")

# Define the replacements
replacements <- list(
  "Very unfav" = 1,
  "Somewhat u" = 2,
  "Somewhat f" = 3,
  "Very favor" = 4
)

# Apply lapply for recoding
favc[columns] <- as.data.frame(lapply(favc[columns], function(column) {
  recode(column, !!!replacements)
}))

favc$roots <- relevel(factor(favc$roots), ref = "Non-Communist")

# count and percentage calculation

favc$pid7 <- as.factor(favc$pid7)


# model for Communist group
favc_communist <- subset(favc, roots == "Communist")
favc_model_communist <- lm(c2 ~ pid7, data = favc_communist)

# model for Non-Communist group
favc_non_communist <- subset(favc, roots == "Non-Communist")
favc_model_non_communist <- lm(c2 ~ pid7, data = favc_non_communist)

# predictions for Communist group
predictions_communist <- predict(favc_model_communist, interval = "confidence")
favc_communist$fit <- predictions_communist[, "fit"]
favc_communist$lwr <- predictions_communist[, "lwr"]
favc_communist$upr <- predictions_communist[, "upr"]

# predictions for Non-Communist group
predictions_non_communist <- predict(favc_model_non_communist, interval = "confidence")
favc_non_communist$fit <- predictions_non_communist[, "fit"]
favc_non_communist$lwr <- predictions_non_communist[, "lwr"]
favc_non_communist$upr <- predictions_non_communist[, "upr"]

# Combined groups
favc_combined <- rbind(favc_communist, favc_non_communist)


# Create the plot
ggplot(favc_combined, aes(x = pid7, y = fit, color = roots)) +
  geom_point(size = 1,  position = position_dodge(width = 0.4)) +  # Points for predicted values
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, position = position_dodge(width = 0.4)) +
  ylim(1, 4.1) +  # Adjust y-axis range
  labs(
    x = "Party Identification from Strong Democrat to Strong Republican", 
    y = "Average Obama Approval \n (ranging from 1 = Very Unfavorable to 4 = Very favorable)", 
    title = "Approval of Obama is almost identical fo Communist and Non-Communist \n within each party identification"
  ) +
  scale_x_discrete(labels = c("SD", "D", "LD", "I", "LR", "R", "SR")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 10),
    legend.title = element_blank(),  # Remove legend title
    legend.position = "top"  # Place legend at top
  ) +
  scale_color_manual(values = c("blue", "red"))  # Customize colors for each group





```

Looking at Obama we can see that the difference in expected approval for each party identification is much closer to being the same for a non-socialist candidate than is the case for Bernie Sanders. It should be noted that although Sanders seems to be less popular than Obama among Latino Americans with roots in Non-Communist countries, they are still more popular than Hillary Clinton among Latinos, regardless of party affiliation or where they trace their ancestry too.

## Conclusion

While this data is insufficient to determine if there is a statistically significant difference between the approval of socialist candidates among Latino Americans who trace their ancestry to Cuba, Venezuala, and Nicaragua, and those who do not, it does show that it should be further investigated. 

By controlling for party affiliation, the difference in approval of the Democratic Socialist Candidate Bernie Sanders is insignificant, however, a larger sample size could narrow the confidence interval to provide clear results. Furthermore, since the survey did not control for age, it is possible that young Americans, who were more supportive of Bernie Sanders in 2016, but do not vote in high proportions made the data more mixed. Subsequent research could add age control or weigh each response by average voter turnout for that age group to make the predictions of favorability more significant and accurate.

It should be noted that while that even if the difference were statistically significant, it is possible for a Socialist Candidate like Bernie Sanders to poll higher with Latino Americans of any ancestry than a non-socialist candidate would. This is exemplified in the appendix by the fact that the average favorability of Sanders was higher for both groups and all party identifications than for Clinton.


## Appendix

```{r, echo = FALSE}
#Clinton Approval
favc <- subset(truncated, c4 != "Not famili")

columns <- c("c4")

# Define the replacements
replacements <- list(
  "Very unfav" = 1,
  "Somewhat u" = 2,
  "Somewhat f" = 3,
  "Very favor" = 4
)

# Apply lapply for recoding
favc[columns] <- as.data.frame(lapply(favc[columns], function(column) {
  recode(column, !!!replacements)
}))

# count and percentage calculation
favc$roots <- relevel(factor(favc$roots), ref = "Non-Communist")

favc$pid7 <- as.factor(favc$pid7)


# model for Communist group
favc_communist <- subset(favc, roots == "Communist")
favc_model_communist <- lm(c4 ~ pid7, data = favc_communist)

# model for Non-Communist group
favc_non_communist <- subset(favc, roots == "Non-Communist")
favc_model_non_communist <- lm(c4 ~ pid7, data = favc_non_communist)

# predictions for Communist group
predictions_communist <- predict(favc_model_communist, interval = "confidence")
favc_communist$fit <- predictions_communist[, "fit"]
favc_communist$lwr <- predictions_communist[, "lwr"]
favc_communist$upr <- predictions_communist[, "upr"]

# predictions for Non-Communist group
predictions_non_communist <- predict(favc_model_non_communist, interval = "confidence")
favc_non_communist$fit <- predictions_non_communist[, "fit"]
favc_non_communist$lwr <- predictions_non_communist[, "lwr"]
favc_non_communist$upr <- predictions_non_communist[, "upr"]

# Combined groups
favc_combined <- rbind(favc_communist, favc_non_communist)


# Create the plot
ggplot(favc_combined, aes(x = pid7, y = fit, color = roots)) +
  geom_point(size = 1,  position = position_dodge(width = 0.4)) +  # Points for predicted values
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, position = position_dodge(width = 0.4)) +
  ylim(1, 4.1) +  # Adjust y-axis range
  labs(
    x = "Party Identification from Strong Democrat to Strong Republican", 
    y = "Average Clinton  Approval \n (ranging from 1 = Very Unfavorable to 4 = Very favorable)", 
    title = "Approval of Clinton is almost identical for Communist and Non-Communist \n within each party identification"
  ) +
  scale_x_discrete(labels = c("SD", "D", "LD", "I", "LR", "R", "SR")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 10),
    legend.title = element_blank(),  # Remove legend title
    legend.position = "top"  # Place legend at top
  ) +
  scale_color_manual(values = c("blue", "red"))  # Customize colors for each group
```

The data suggests that although Hillary Clinton polls at roughly the same amount with Latinos from all poltical identity responses in contrast to Bernie Sanders, Sanders still has a higher average approval from all political party affiliations. 


```{r, echo = FALSE}

#bootstrap


set.seed(1)
n_bootstrap <- 5000

# Bootstrap function to get predictions
bootstrap_predictions <- function(data, n_bootstrap) {
  # Create a container to store predictions
  prediction_list <- vector("list", n_bootstrap)
  
  for (i in 1:n_bootstrap) {
    # Resample data
    resampled_data <- data[sample(1:nrow(data), replace = TRUE), ]
    
    # Fit the model
    b_model <- lm(c8 ~ pid7, data = resampled_data)
    
    # Predict for all levels of pid7
    prediction_list[[i]] <- predict(
      b_model, 
      newdata = data.frame(pid7 = levels(data$pid7))
    )
  }
  
  # Combine predictions into a matrix
  prediction_matrix <- do.call(cbind, prediction_list)
  
  # Calculate summary statistics
  summary_stats <- apply(prediction_matrix, 1, function(x) {
    c(mean = mean(x), lwr = quantile(x, 0.025), upr = quantile(x, 0.975))
  })
  
  # Return as a data frame
  summary_df <- as.data.frame(t(summary_stats))
  summary_df$pid7 <- levels(data$pid7)  # Add pid7 levels
  return(summary_df)
}

# Run bootstrap for Communist and Non-Communist groups


summary_communist <- bootstrap_predictions(fav_communist, n_bootstrap)
summary_communist$roots <- "Communist"



summary_non_communist <- bootstrap_predictions(fav_non_communist, n_bootstrap)
summary_non_communist$roots <- "Non-Communist"


# Combine summaries
combined <- rbind(summary_non_communist, summary_communist)

# Convert pid7 to factor
combined$pid7 <- factor(combined$pid7)
combined$roots <- relevel(factor(combined$roots), ref = "Non-Communist")

# Rename columns in the combined data frame
colnames(combined)[colnames(combined) == "lwr.2.5%"] <- "lwr"
colnames(combined)[colnames(combined) == "upr.97.5%"] <- "upr"



```

```{r, echo = FALSE}

# Plot using bootstrap data
ggplot(combined, aes(x = pid7, y = mean, color = roots)) +
  geom_point(size = 1, position = position_dodge(width = 0.4)) +  # Points for predicted values
  geom_errorbar(aes(ymin = lwr, ymax = upr), width = 0.2, position = position_dodge(width = 0.4)) +
  ylim(1, 4) +  # Adjust y-axis range
  labs(
    x = "Party Identification from Strong Democrat to Strong Republican", 
    y = "Average Sanders Approval \n (ranging from 1 = Very Unfavorable to 4 = Very favorable)", 
    title = "Approval of Sanders does not vary between Communist and non-Communist \n to a statistically significant extent within each party identification group"
  ) +
  scale_x_discrete(labels = c("SD", "D", "LD", "I", "LR", "R", "SR")) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(size = 12),
    axis.title.x = element_text(size = 14),
    axis.title.y = element_text(size = 10),
    legend.title = element_blank(),  # Remove legend title
    legend.position = "top"  # Place legend at top
  ) +
  scale_color_manual(values = c("blue", "red"))  # Customize colors for each group


```

While bootstrapping does narrow the confidence interval, the findings are still statistically insignificant.


```{r, echo = FALSE}
# Calculating effect of bootstrap
combined$ci <- combined$upr - combined$lwr
fav_combined$ci <- fav_combined$upr - fav_combined$lwr
# Merge based on pid7 and roots
comparison <- merge(
  combined[, c("pid7", "roots", "ci")], 
  fav_combined[, c("pid7", "roots", "ci")], 
  by = c("pid7", "roots"), 
  suffixes = c("_bootstrap", "_original")
)

comparison$ci_change <- (1 -comparison$ci_bootstrap/comparison$ci_original)*100

unique_comparison <- comparison[!duplicated(comparison$ci_bootstrap), ]

unique_comparison

```

The bootstrap does not have a uniform effect on confidence interval for the predicted favorability.




